Iterative forecasting and the trap of self-fulfilling prophecies

### 2021-04-02 manuscript/appendix.Rmd
tidy :broom:

---
title: 'Appendix'
date: "`r Sys.Date()`"
author: Carl Boettiger
journal: TBD
layout: 3p
bibliography: refs.bib
#output:
#  tufte::tufte_handout:
#    dev: cairo_pdf
#    latex_engine: xelatex
output: 
  hrbrthemes::ipsum_pdf:
    dev: cairo_pdf
    latex_engine: xelatex

@@

## Optimal quotas as a Markov Decision Process (MDP) problem

@@

# Ecological Models

@@

where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standar-deviation $\sigma_i$.

@@

Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.  However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].  As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.  While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous. Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].

Below, I provide annotated code necessary to completely reproduce all of the analysis presented in the main paper. This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse]. This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.  The code below generates the data tables required for each plot shown in the main paper.  Data are stored as `.csv` tables to be accessible to any relevant software program. Plotting commands used in the main paper are embedded in the paper's RMarkdown file at <https://github.com/cboettig/bad-forecast-good-decision>.

@@

# Model definitions

@@

# Optimal control solutions

@@

It is straight forward to apply the proper scoring formula of @Gneiting2007 
based on the first two moments of the distribution to score the respective
forecasts under both the unfished and actively managed scenario sfor each model:

@@

# Posteriors

>>>>

---
title: 'Appendix'
date: "`r Sys.Date()`"
author: Carl Boettiger
journal: TBD
layout: 3p
bibliography: refs.bib
csl: csl/ecology-letters.csl
#output:
#  tufte::tufte_handout:
#    dev: cairo_pdf
#    latex_engine: xelatex
output: 
  hrbrthemes::ipsum_pdf:
    dev: cairo_pdf
    latex_engine: xelatex

header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers

@@

Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.  However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulas or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].  As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.  While computer code can be more verbose and difficult to read than mathematical formulas presented here, it is also less ambiguous. Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in MatLab [@Marescot2013; @MDPtoolbox].

Below, I provide carefully annotated code necessary to completely reproduce all of the analysis presented in the main paper. This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse]. This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.  The code below generates the data tables required for each plot shown in the main paper.  Data are stored as `.csv` tables to be accessible to any relevant software program. Plotting commands used in the main paper are embedded in the paper's RMarkdown file at <https://github.com/cboettig/bad-forecast-good-decision>.

# Optimal quotas as a Markov Decision Process (MDP) problem

@@

## Ecological Model Definitions

@@

where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.

@@

@@

@@

## Optimal control solutions

@@

It is straight forward to apply the proper scoring formula of @Gneiting2007 
based on the first two moments of the distribution to score the respective
forecasts under both the un-fished and actively managed scenario for each model:

@@

## Posteriors

### 2021-04-02 manuscript/manuscript.Rmd
tidy :broom:

## change to true to add optional line numbering
lineno: true
csl: ecology-letters.csl
output: rticles::elsevier_article
layout: 3p
journal: TBD
---

@@

Iterative learning approaches such as adaptive management or iterative forecasting [@Dietze2018] are particularly compelling because they address a different reason why seemingly-accurate models can lead to poor outcomes: the problem of over-fitting.
When models are estimated from the data they seek to describe, the model may over-fit the data, making the data seem even more likely than it would be under the true process [@Burnham1998].
Over-fit models may underestimate uncertainty and make poor predictions about future outcomes [@Ginzburg2004].
Because any iterative learning compares models to future outcomes, overfitting those outcomes is impossible.
This concept was rigorously formalized by @Gneiting2007's proof of "proper" scoring rules.
By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on average, than that of the underlying process, $P(x)$.
@Gneiting2007 further shows which techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which cannot be over-fit.
Adaptive management is based on equally secure footing because model predictions are again compared to future outcomes not used to fit the model.
Instead of scoring and selecting among alternative models, adaptive management considers probabilities over all models [@Walters1978].
These two approaches are particularly compelling because they can both reduce uncertainty and avoid over-fitting.
My example is not meant to undercut the importance of these approaches.
Previous work has long acknowledged the panelope of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives implementation errors, or lack of resources for monitoring and updating [e.g. @Ludwig1993].
Here, the problem is more fundamental.

@@

Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does little to resolve this problem [Fig 3B]. 
Iterative updates again quickly dismiss the parameter values assumed by model 1, though with more options to choose from, this probability is spead over a range of seemingly plausible candidate models instead of a single alternative model (see Appendix).
While the adaptive management of additional actions and observations slowly narrow this subset of plausible models, decisions based on this uncertainty prevent fish stocks from recovering fully, and realize lower harvests as a result.
Note that learning under either adaptive management approach (using two models or 42), the decision-maker becomes ever-increasingly convinced that they are using the right model or models.
Future stock sizes fall consistently in the range predicted by the model(s), and consistently outside the range predicted by model 1.
Consequently, each iteration the managers are only more firmly convinced that they are maintaining the fish stock near the biomass that supports maximum sustainable yield, when in fact they are sustaining a harvest regime that is preventing recovery of the stock to the much higher productivity regime which would have been achieved under model 1.

@@

<!-- Intuition for it's not just this example -->
This phenomenon is not unique to sequential decision problems or optimal control solutions.
For example, Management Strategy Evaluation [MSE; @Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
That approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
Doing so, we would find the best constant-mortality solution under model 1 performs much worse than the best constant-mortality solution under model 2.
Does this mean MSE is not susceptible to this issue?
No, it does not.
It is just as easy to construct an alternative model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
If model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from model 1 and towards worse-performing models once again.
Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar model 1.
This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the havest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.

@@

Worse, absent the counterfactual of non-adaptive management to compare to, this detioriation in outcomes is invisible. 
All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
By responding in any way to future observations to reduce model uncertainty, our manager is caught in a trap of self-fulfilling prophecy:
believing ever more firmly in models that make more accurate predictions while driving decisions that prevent the fish stock from recovering.
The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
It may seem that the adaptive management approach fails because posterior probabilities are updated according to Bayes rule.
Like iterative forecasting, this favors models which better predict the data, despite the fact that under the adaptive management approach, the overall optimization is conditioned on actual utility and not model fit.
However, this issue is not easily avoided.
For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we deterime that model 1 leads to better outcomes.
This raises several questions.
Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? If so, How would we know, and what should we do about it?

>>>>

header-includes:
  - \linenumbers

csl: csl/ecology-letters.csl
output: rticles::elsevier_article
layout: 3p
journal: TBD
---

@@

Iterative learning approaches such as adaptive management or iterative forecasting [@Dietze2018] are particularly compelling because they address a different reason why seemingly-accurate models can lead to poor outcomes: the problem of over-fitting.
When models are estimated from the data they seek to describe, the model may over-fit the data, making the data seem even more likely than it would be under the true process [@Burnham1998].
Over-fit models may underestimate uncertainty and make poor predictions about future outcomes [@Ginzburg2004].
Because any iterative learning compares models to future outcomes, over-fitting those outcomes is impossible.
This concept was rigorously formalized by @Gneiting2007's proof of "proper" scoring rules.
By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on average, than that of the underlying process, $P(x)$.
@Gneiting2007 further shows which techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which cannot be over-fit.
Adaptive management is based on equally secure footing because model predictions are again compared to future outcomes not used to fit the model.
Instead of scoring and selecting among alternative models, adaptive management considers probabilities over all models [@Walters1978].
These two approaches are particularly compelling because they can both reduce uncertainty and avoid over-fitting.
My example is not meant to undercut the importance of these approaches.
Previous work has long acknowledged the panoply of ways in which iterative learning or other model-based decision making can go astray due to conflicting incentives implementation errors, or lack of resources for monitoring and updating [e.g. @Ludwig1993].
Here, the problem is more fundamental.

@@

Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does little to resolve this problem [Fig 3B]. 
Iterative updates again quickly dismiss the parameter values assumed by model 1, though with more options to choose from, this probability is spread over a range of seemingly plausible candidate models instead of a single alternative model (see Appendix).
While the adaptive management of additional actions and observations slowly narrow this subset of plausible models, decisions based on this uncertainty prevent fish stocks from recovering fully, and realize lower harvests as a result.
Note that learning under either adaptive management approach (using two models or 42), the decision-maker becomes ever-increasingly convinced that they are using the right model or models.
Future stock sizes fall consistently in the range predicted by the model(s), and consistently outside the range predicted by model 1.
Consequently, each iteration the managers are only more firmly convinced that they are maintaining the fish stock near the biomass that supports maximum sustainable yield, when in fact they are sustaining a harvest regime that is preventing recovery of the stock to the much higher productivity regime which would have been achieved under model 1.

@@

<!-- Intuition for it's not just this example -->
This phenomenon is not unique to sequential decision problems or optimal control solutions.
For example, Management Strategy Evaluation [MSE; @Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.
That approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.
Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem.
Doing so, we would find the best constant-mortality solution under model 1 performs much worse than the best constant-mortality solution under model 2.
Does this mean MSE is not susceptible to this issue?
No, it does not.
It is just as easy to construct an alternative model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.
Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954].
If model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from model 1 and towards worse-performing models once again.
Many optimization problems share this feature in which the optimal policy depends only on a subset or ratio of model parameters, such that it is usually easy to find a similar model 1.
This example also illustrates the importance of decision constraints -- a model that gives very good outcomes when we are free to vary the harvest quota each year (the optimal control problem here) may give very poor results under the constraint of constant mortality, and vice versa.

@@

Worse, absent the counter-factual of non-adaptive management to compare to, this deterioration in outcomes is invisible. 
All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
By responding in any way to future observations to reduce model uncertainty, our manager is caught in a trap of self-fulfilling prophecy:
believing ever more firmly in models that make more accurate predictions while driving decisions that prevent the fish stock from recovering.
The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
It may seem that the adaptive management approach fails because posterior probabilities are updated according to Bayes rule.
Like iterative forecasting, this favors models which better predict the data, despite the fact that under the adaptive management approach, the overall optimization is conditioned on actual utility and not model fit.
However, this issue is not easily avoided.
For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we determine that model 1 leads to better outcomes.
This raises several questions.
Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? If so, How would we know, and what should we do about it?
